Toward Resolved Simulations of Burning Fronts in Thermonuclear X-ray Bursts

### 2018-09-26 astronum_2018/astronum-xrb.tex
Add IH affiliation.
Minor rephrasing or restructuring.

Global shallow water simulations~\cite{SPIT_ETAL02} capture the
large scale spreading of the burning front by using a very simple
vertical structure.  These simulations were instrumental in showing
that the Coriolis force plays an important role in confining a
spreading region and the flame spreading can explain brightness
oscillations during the rise of the burst lightcurve.

@@

Flame spreading was first modeled in a series of calculations
employing an algorithm from atmospheric science, where vertical
hydrostatic equilibrium is enforced and wide-aspect ratio zones give a
horizontal CFL number that is large enough to obtain long timescale
evolution~\cite{cavecchi:2012}.

These calculations display key features of the propagation
mechanism, namely a balance between hydrodynamics and the pure
flame physics. The flame proceeds mainly via conduction, thus being a
deflagration. The flame front, however, is not vertical but inclined,
the angle with the horizontal being very small: $\theta \sim
H/R_{\rm{Ro}} \sim 10^{-3}$. Here $H$ is the scale height and
$R_{\rm{Ro}}$ is the Rossby radius. It is the hydrodynamical balance
between the Coriolis force, gravity and pressure gradients that
determines the Rossby radius and therefore the inclination angle of the
flame front. The extended surface of the flame front is what leads to
propagation speeds of order $10^5 $~km/s, in good agreement with
observations. Due to its dependence on the Coriolis force, the flame
speed changes across the surface of the NS (being faster at the
equator), leading to potentially detectable effects if the rise of the
burst is observed with enough resolution
\cite{art-2015-cavecchi-etal}. The NS systems exhibiting XRBs are
expected to have magnetic fields in the range $10^7 - 10^{10}$~G.
Inclusion of a vertical magnetic field proved to have a strong
influence on the nature and the speed of flame propagation
\cite{art-2016-cavecchi-etal}, due to the interaction between the
Coriolis force and the magnetic tension, which further acts towards
determining the inclination angle of the flame front.

A longstanding goal is to perform simulations where we resolve the
flame structure, to allow us to accurately capture the
nucleosynthesis, and watch it move across the neutron star surface.
The results of these simulations will enable strict comparison to
observations made with the new generation of X-ray telescopes such as
\textit{eXTP} and \textit{STROBE--X}
\cite{art-2017-wilhod-etal,art-2016-zhang-etal} which will offer
high collecting area and time resolution.
We show some in-progress calculations that are a step toward this (but
still have approximations) and also discuss what new algorithmic
techniques might be needed to realize this goal with the advent of
exascale computing.

@@

\begin{figure}
\centering
\includegraphics[width=0.75\linewidth]{xrb}
\caption{\label{fig:xrb} Helium flame spreading across the surface of
  a neutron star.  Shown are temperature (top), the velocity out of the
  simulation plane (middle), and nuclear energy generation rate
  (bottom).
  The out of plane velocity contours indicate where the Coriolis force
  has created a hurricane-like wind structure which confines the flame,
  leading to an inclined front. It is at such interface that the
  temperature and burning rate are highest.
  }
\end{figure}

>>>>

Global shallow water simulations~\cite{SPIT_ETAL02} capture the
large scale spreading of the burning front by using a very simple
vertical structure.  These simulations were instrumental in showing
that the Coriolis force plays an important role in confining a
spreading region and that the flame spreading can explain brightness
oscillations during the rise of the burst lightcurve.

@@

Flame spreading was first modeled in a series of calculations
employing an algorithm from atmospheric science, where vertical
hydrostatic equilibrium is enforced and wide-aspect ratio zones give a
horizontal CFL number that is large enough to obtain long timescale
evolution~\cite{cavecchi:2012}.
%
These calculations display key features of the propagation
mechanism, namely a balance between hydrodynamics and the pure
flame physics. The flame proceeds mainly via conduction, thus being a
deflagration. The flame front, however, is not vertical but inclined,
the angle with the horizontal being very small: $\theta \sim
H/R_{\rm{Ro}} \sim 10^{-3}$. Here $H$ is the scale height and
$R_{\rm{Ro}}$ is the Rossby radius. It is the hydrodynamical balance
between the Coriolis force, gravity and pressure gradients that
determines the Rossby radius and therefore the inclination angle of the
flame front. The extended surface of the flame front is what leads to
propagation speeds of order $10^5 $~km/s, in good agreement with
observations. Due to its dependence on the Coriolis force, the flame
speed changes across the surface of the NS (being faster at the
equator), leading to potentially detectable effects if the rise of the
burst is observed with enough resolution
\cite{art-2015-cavecchi-etal}.

The NS systems exhibiting XRBs are
expected to have magnetic fields in the range $10^7 - 10^{10}$~G.
Inclusion of a vertical magnetic field proved to have a strong
influence on the nature and the speed of flame propagation
\cite{art-2016-cavecchi-etal}, due to the interaction between the
Coriolis force and the magnetic tension, which further acts towards
determining the inclination angle of the flame front.

A longstanding goal is to perform simulations where we resolve the
flame structure, to allow us to accurately capture the
nucleosynthesis, and watch it move across the neutron star surface.
The results of these simulations will enable strict comparison to
observations made with the new generation of X-ray telescopes such as
\textit{eXTP} and \textit{STROBE--X}
\cite{art-2017-wilhod-etal,art-2016-zhang-etal} which will offer
high collecting area and time resolution.
We show some in-progress calculations that are a step toward this, discuss
their remaining approximations, and also discuss what new algorithmic
techniques might be needed to realize this goal with the advent of
exascale computing.

@@

\begin{figure}
\centering
\includegraphics[width=0.75\linewidth]{xrb}
\caption{\label{fig:xrb} Helium flame spreading across the surface of
  a neutron star.  Shown are temperature (top), the velocity out of the
  simulation plane (middle), and nuclear energy generation rate
  (bottom).
  The out of plane velocity contours indicate where the Coriolis force
  has created a hurricane-like wind structure which confines the flame,
  leading to an inclined front. It is at such an interface that the
  temperature and burning rate are highest.
  }
\end{figure}
